---
title: "Game Day Predictions — March 29, 2026"
date: "March 29, 2026"
categories: [predictions, MLB, machine learning]
---

Today is the first day we're running our ML pipeline live. The models were 
trained on 2023–2024 MLB data and this is their debut. We're publishing the 
raw output and will score everything tomorrow once results are in.

A few things to keep in mind as you read:

- It's opening weekend, so most teams have played fewer than 5 games. The 
  rolling averages are essentially priors from last season — the models will 
  get sharper as the sample grows.
- Win probability and predicted totals are the two numbers we care about most. 
  The run line lean only triggers when win probability exceeds 65%, which 
  won't happen often.
- Not all games are present, we are still working on the pipeline
- This is predictive modeling, not sports betting

Additionally, do not expect these to occur every day.

---

## Kansas City Royals @ Atlanta Braves
```{r}
#| echo: false
cat(readLines("reports/2026-03-29/KC_at_ATL_2026-03-29.txt"), sep = "\n")
```

---

## Minnesota Twins @ Balitmore Orioles
```{r}
#| echo: false
cat(readLines("reports/2026-03-29/MIN_at_BAL_2026-03-29.txt"), sep = "\n")
```

---

## Texas Rangers @ Philadelphia Phillies
```{r}
#| echo: false
cat(readLines("reports/2026-03-29/TEX_at_PHI_2026-03-29.txt"), sep = "\n")
```

---

## Colorado Rockies @ Miami Marlins
```{r}
#| echo: false
cat(readLines("reports/2026-03-29/COL_at_MIA_2026-03-29.txt"), sep = "\n")
```

---

## Tampa Bay Rays @ St. Louis Cardinals
```{r}
#| echo: false
cat(readLines("reports/2026-03-29/TB_at_STL_2026-03-29.txt"), sep = "\n")
```

---

## Cleveland Guardians @ Seattle Mariners
```{r}
#| echo: false
cat(readLines("reports/2026-03-29/CLE_at_SEA_2026-03-29.txt"), sep = "\n")
```

---

*Predictions generated by ProPlotFits ML pipeline. Models trained on 
2023–2024 MLB regular season data. First live prediction season — 
we will score these tomorrow.*